Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-01-06 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Added a couple paragraphs to section 2 and refined the simulations some.  also moved the section of the paper specific rmd commands to the end

The recent (and on-going) COVID-19 global pandemic has galvanized public interest in understanding more about infectious disease modeling and has highlighted the usefulness of research in the area of infectious disease epidemiology. Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly. Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g. flue prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g. ...][]{}^[\@Shannon: CITE]. At the same time, descriptive statistics and visualizations (from universities like John Hopkins \citep{}^[\@Shannon: CITE], many branches and levels of government, and new organizations \citep{}^[\@Shannon: CITE]) are an important first step of the process.

@@

Although epidemics are by definition a process that evolves over time, epidemics, including the recent COVID-19, often need to be compared in a time-invariant way to understand the processes at play. Additionally, many tools to examine the quantity of the population in each state along the infection process (for example: quantity of suspectible vs infected vs recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states. Tools in \pkg{EpiCompare} attempt to give the user the ability to extend their toolkit to evaluate epidemics to also include time-invariant approaches and in this section we present benefits of the time-invariant analysis, with 1) motivation for time-invariant analysis through $R_0$ (the initial reproduction number), 2) time-invariant visualization tools for 3-state models (e.g. SIR models), and 3) potential for similar analysis for models with more states (e.g. SEIR models). \ben{Will need to rewrite this section "introduction".}

@@

$R_0$, the initial reproduction number, has been called the "most important quantity in epdiemiology" \citep[][]{heesterbeek2003}. By definition, $R_0$ is the number of expected secondary infections when a primary infection is introduced to a susceptible population. To see the importance of $R_0$, one need only read the newspaper \citep{Fisher2020} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{Aronson2020}. $R_0$ is also, maybe, the most famous \textit{time-invariant} numerical summary of an epidemic, which allows epidemics to be compared to one another in both different time and geographic scales.

@@

In regards to traditional $state$ vs. time plots, $R_0$ is difficult to visualize, especially with respect from one epidemic to another. For example, consider the scenarios where the first epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_1 = .3$ and $\gamma_1 = .15$, and the second epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_2 = .24$ and $\gamma_1 = .12$ over 40 days.  Both epidemics have the same value of $R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2$.  The epidemic trajectories are shown in the $state$ vs. time plots in Figure \ref{fig:sir-ex}.  At a glance, we may assume that Model 1 has a larger $R_0$ than Model 2 because the peak of infection occurs more quickly than in Model 2.  On the other hand, we may think Model 2 has a larger $R_0$ because it has a slightly larger peak of infection than Model 1.

@@

```{r echo = FALSE, fig.cap = "\\label{fig:sir-ex}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$."  }
set.seed(1225)

@@

ggplot(data = df_long,
       aes(x = t, y = value, col = type)) + 
  facet_wrap(~name, ncol = 1, scales = "free") +
  geom_point() +
  scale_color_manual(values = c("red", "blue"), name = "Epidemic") +
  labs(x = "Time",
       y = "# in state",
       title = latex2exp::TeX("Comparing two epidemics over 50 days with same $R_0$"))
```
However, when we plot the trajectories as a single curve using the ternary plot in the time-invariant view, we immediately see a different story.  In this time-invariant view in Fig. \ref{fig:ternary-ex}, the points seem to overlap and form the same trajectory.  In other words, we can see there is something fundamentally linking these two different epidemics, and this fundamental link turns out to be $R_0$.

@@

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Appendix \ref{app:proof}.  The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model is also a point on the epidemic path of the second SIR model. Taking the sample simulations from Fig. \ref{fig:different-scales-standard}, Fig. \ref{fig:different-scales-tern} presents these models in a ternary plot.

```{r warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "\\label{fig:ternary-ex}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.  These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered."}
ggplot(data = df) +
  geom_point(aes(x = S, y = I, z = R,
                     col = type)) +
    coord_tern() + 
  theme_sir() +
  scale_color_manual(values = c("red", "blue"), name = "Epidemic") 
```

## Simplices beyond ternary plots and time-invariant comparison tools {short-title="Simplexes beyond ternary plots" #sec:time-invariant-in-simplex}

Beyond ternary plots, and the associated SIR models, we can extend the time-invariant approach described in Section \ref{sec:r0} to higher dimensions (i.e. models with more states). The constraints in 3d that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a spaces of 3d simplexes, and the ternary plot specifically represents these after scaling to examine such values as proportions (ternary plots are known as a 3d unit simplex due to it's scaling). This same scaling for larger models (i.e. with more states) can be done onto different simplexes. In this package we present tools to help compare models (mostly through similations), and this tool compares these objects after projecting them into a one-dimension less space through the simplexical structure of the data.

In \pkg{EpiCompare} we allow for the compare epidemics in these higher dimensional spaces by first projecting onto these simplexes. Even though higher dimensional models may not be able to visualized, we provide multiple tools to aid in the comparison of models and epidemics. The first of which uses multiple simulations under specific model parameters to assess the bairabilty of the model fit. In \pkg{EpiCompare} we provide ways to create prediction regions for a true epidemic under the model assumptions using these simulations. These regions require representing multi-dimensional structures for functions to completely contain epidemics, and treat these simulations and epidmeics as filamental objects. We extend off of papers like \citet{Dalmasso2019a} to create these bands. These high dimensional bands allow the user to assess if the true epidemic is within the band (there-by assessing the model's representation of the epidemic), and also compare who different models are from each other through distances that compare sets. We recommend using the Hausdorff distance to compare such sets as it captures how much bigger the sets would have to expand to cover each other, and is defined mathematically as 
\[
d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y) \;,\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
\]

@@

```{r echo = FALSE, results = 'hide'}
with(hagelloch_raw, sum(PRO > ERU, na.rm = TRUE))

>>>>

The recent (and on-going) COVID-19 global pandemic has galvanized public interest in understanding more about infectious disease modeling and has highlighted the usefulness of research in the area of infectious disease epidemiology. Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly. Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g. flu prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g. ...][]{}^[\@Shannon: CITE]. At the same time, descriptive statistics and visualizations (from universities like John Hopkins \citep{}^[\@Shannon: CITE], many branches and levels of government, and new organizations \citep{}^[\@Shannon: CITE]) are an important first step of the process.

@@

Although epidemics are by definition a process that evolves over time, epidemics, including the recent COVID-19, often need to be compared in a time-invariant way to understand the processes at play. Additionally, many tools to examine the quantity of the population in each state along the infection process (for example: quantity of susceptible vs infected vs recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states. Tools in \pkg{EpiCompare} attempt to give the user the ability to extend their toolkit to evaluate epidemics to also include time-invariant approaches and in this section we present benefits of the time-invariant analysis, with 1) motivation for time-invariant analysis through $R_0$ (the initial reproduction number), 2) time-invariant visualization tools for 3-state models (e.g. SIR models), and 3) potential for similar analysis for models with more states (e.g. SEIR models). \ben{Will need to rewrite this section "introduction".}

@@

 By definition, $R_0$ is the number of expected secondary infections when a primary infection is introduced to a susceptible population.  $R_0$ is also, maybe, the most famous \textit{time-invariant} numerical summary of an epidemic, which allows epidemics to be compared to one another in both different time and geographic scales. For example, $R_0$ for Covid-19 is estimated to be between 2-3, seasonal influenza between 1.2-2, and modern measles outbreaks between 5-6.

@@

With regards to traditional epidemic $state$ vs. $time$ plots, $R_0$ is difficult to visualize, especially with respect from one epidemic to another. For example, consider the scenarios where the first epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_1 = 0.3$ and $\gamma_1 = 0.15$, and the second epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_2 = 0.24$ and $\gamma_1 = 0.12$ over 40 days.  Both epidemics have the same value of $R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2$.  The epidemic trajectories are shown in the $state$ vs. time plots in Figure \ref{fig:sir-ex}.  At a glance, we may assume that Model 1 has a larger $R_0$ than Model 2 because the peak of infection occurs more quickly than in Model 2.  On the other hand, we may think Model 2 has a larger $R_0$ because it has a slightly larger peak of infection than Model 1.

@@

```{r echo = FALSE, fig.cap = "\\label{fig:different-scales-standard}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$."  }
set.seed(1225)
theme_set(theme_bw(base_size = 24))

@@

ggplot(data = df_long,
       aes(x = t, y = value, col = type)) + 
  facet_wrap(~name, ncol = 1, scales = "free") +
  geom_point() +
  scale_color_manual(values = c("red", "blue"), name = "Epidemic") +
  labs(x = "Time",
       y = "# in state",
       title = latex2exp::TeX("Comparing two epidemics over first 15 days with same $R_0$")) +
  coord_cartesian(xlim = c(0, 15))
```
However, when we plot the trajectories as a single curve using the ternary plot in the time-invariant view, we immediately see a different story.  In this time-invariant view in Fig. \ref{fig:ternary-ex}, the points seem to overlap and form the same trajectory.  In other words, we can see there is something fundamentally linking these two different epidemics, and this fundamental link turns out to be $R_0$.

@@

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Appendix \ref{app:proof}.  The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model is also a point on the epidemic path of the second SIR model. Taking the sample simulations from Fig. \ref{fig:different-scales-standard}, Fig. \ref{fig:different-scales-tern} presents these two models in a ternary plot.  This means, that at a glance, we can tell if two epidemics have different values of $R_0$.

```{r warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "\\label{fig:different-scales-tern}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.  These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered."}
ggplot(data = df %>%
         filter(t <= 15)) +
  geom_point(aes(x = S, y = I, z = R,
                     col = type)) +
    coord_tern() + 
  theme_sir() +
  scale_color_manual(values = c("red", "blue"), name = "Epidemic") 
```

## Beyond the Kermack and McKendrick SIR models

Although the result of Theorem \ref{thm:sir-scale} allows for easy visual comparison of $R_0$ in Kermack SIR models, it does require stringent assumptions of homogeneity of behavior in populations.  The use of visualizing epidemics in a time-invariant lens via ternary plots extends beyond those of models that follow the ODEs in the Kermack-McKendrick equations.  Any model with S, I, and R states can be visualized with ternary plots, regardless of birth and death dynamics and regardless of homogeneity of individual behavior.  We can use ternary plots to compare the spread of a disease for groups within a population without time as a confounding factor.  We show an example of this in a later section.

But time-invariant analysis is also useful for epidemic models with more than three states. The constraints in three dimensions that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a space of 3d simplices, and the ternary plot specifically represents these simplices, after scaling to examine such values as proportions between 0 and 1 (ternary plots are known as a 3d unit simplex due to it's scaling). This same scaling for larger models (i.e. with more states) can be done onto different simplexes. In this package we present tools to help compare models (mostly through simulations).  s tool compares these objects after projecting them into a one-dimension-fewer space through the simplexical structure of the data.

In \pkg{EpiCompare} we allow for the comparison epidemics in these higher dimensional spaces by first projecting onto these simplexes. Even though higher dimensional models may not be able to visualized, we provide multiple tools to aid in the comparison of models and epidemics. The first of which uses multiple simulations under specific model parameters to assess the bairabilty of the model fit. In \pkg{EpiCompare} we provide ways to create prediction regions for a true epidemic under the model assumptions using these simulations. These regions require representing multi-dimensional structures for functions to completely contain epidemics, and treat these simulations and epidmeics as filamental objects. We extend off of papers like \citet{Dalmasso2019a} to create these bands. These high dimensional bands allow the user to assess if the true epidemic is within the band (there-by assessing the model's representation of the epidemic), and also compare who different models are from each other through distances that compare sets. We recommend using the Hausdorff distance to compare such sets as it captures how much bigger the sets would have to expand to cover each other, and is defined mathematically as 
\[
d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y) \;,\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
\]

@@

```{r echo = FALSE, results = 'hide'}
data(hagelloch_raw)
with(hagelloch_raw, sum(PRO > ERU, na.rm = TRUE))

### 2021-01-06 jss/EpiCompare-paper/EpiCompare-paper.tex
Added a couple paragraphs to section 2 and refined the simulations some.  also moved the section of the paper specific rmd commands to the end

The recent (and on-going) COVID-19 global pandemic has galvanized public
interest in understanding more about infectious disease modeling and has
highlighted the usefulness of research in the area of infectious disease
epidemiology. Infectious diseases inflict enormous burdens on the world:
millions of lives lost and trillions of dollars spent yearly. Infectious
disease models typically attempt to do one or more of the following: 1)
predict the spread of current and future epidemics
\citep[e.g. flue prediction][]{Biggerstaff2016}, 2) analyze past and
current epidemics to increase scientific knowledge
\citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or
project epidemic scenarios under pre-specified parameters
\citep[e.g. ...][]{}\footnote{@Shannon: CITE}. At the same time,
descriptive statistics and visualizations (from universities like John
Hopkins \citep{}\footnote{@Shannon: CITE}, many branches and levels of
government, and new organizations \citep{}\footnote{@Shannon: CITE}) are
an important first step of the process.

@@

Although epidemics are by definition a process that evolves over time,
epidemics, including the recent COVID-19, often need to be compared in a
time-invariant way to understand the processes at play. Additionally,
many tools to examine the quantity of the population in each state along
the infection process (for example: quantity of suspectible vs infected
vs recovered individuals) don't always as intelligently capture the
natural connections between the proportion of individuals in these
states. Tools in \pkg{EpiCompare} attempt to give the user the ability
to extend their toolkit to evaluate epidemics to also include
time-invariant approaches and in this section we present benefits of the
time-invariant analysis, with 1) motivation for time-invariant analysis
through \(R_0\) (the initial reproduction number), 2) time-invariant
visualization tools for 3-state models (e.g.~SIR models), and 3)
potential for similar analysis for models with more states (e.g.~SEIR
models).
\textcolor{violet}{Will need to rewrite this section "introduction".}

@@

\(R_0\), the initial reproduction number, has been called the ``most
important quantity in epdiemiology'' \citep[][]{heesterbeek2003}. By
definition, \(R_0\) is the number of expected secondary infections when
a primary infection is introduced to a susceptible population. To see
the importance of \(R_0\), one need only read the newspaper
\citep{Fisher2020} or look at the length of the table of estimated
\(R_0\) quantities for COVID-19 \citep{Aronson2020}. \(R_0\) is also,
maybe, the most famous \textit{time-invariant} numerical summary of an
epidemic, which allows epidemics to be compared to one another in both
different time and geographic scales.

@@

In regards to traditional \(state\) vs.~time plots, \(R_0\) is difficult
to visualize, especially with respect from one epidemic to another. For
example, consider the scenarios where the first epidemic is generated
from a SIR model with \((S(0) = 990, I(0) = 10)\), \(\beta_1 = .3\) and
\(\gamma_1 = .15\), and the second epidemic is generated from a SIR
model with \((S(0) = 990, I(0) = 10)\), \(\beta_2 = .24\) and
\(\gamma_1 = .12\) over 40 days. Both epidemics have the same value of
\(R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2\). The epidemic
trajectories are shown in the \(state\) vs.~time plots in Figure
\ref{fig:sir-ex}. At a glance, we may assume that Model 1 has a larger
\(R_0\) than Model 2 because the peak of infection occurs more quickly
than in Model 2. On the other hand, we may think Model 2 has a larger
\(R_0\) because it has a slightly larger peak of infection than Model 1.

@@

\caption{\label{fig:sir-ex}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.}\label{fig:unnamed-chunk-2}
\end{figure}
\end{CodeChunk}

@@

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
result from \cite{Harko2014} and is shown in detail in Appendix
\ref{app:proof}. The consequence of Theorem \ref{thm:sir-scale} is that
for two SIR models that have the same initial percent of individuals in
each state and \(R_0\) then for every point on the epidemic path of the
first SIR model is also a point on the epidemic path of the second SIR
model. Taking the sample simulations from Fig.
\ref{fig:different-scales-standard}, Fig.
\ref{fig:different-scales-tern} presents these models in a ternary plot.

@@

\caption{\label{fig:ternary-ex}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.  These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered.}\label{fig:unnamed-chunk-3}
\end{figure}
\end{CodeChunk}

\subsection[Simplexes beyond ternary plots]{Simplices beyond ternary
plots and time-invariant comparison
tools}\label{sec:time-invariant-in-simplex}

Beyond ternary plots, and the associated SIR models, we can extend the
time-invariant approach described in Section \ref{sec:r0} to higher
dimensions (i.e.~models with more states). The constraints in 3d that
are met with the SIR model (that is
\(\sum_{i=1}^3 (\text{number in state}(i)) = N(t)\)) actually represents
a spaces of 3d simplexes, and the ternary plot specifically represents
these after scaling to examine such values as proportions (ternary plots
are known as a 3d unit simplex due to it's scaling). This same scaling
for larger models (i.e.~with more states) can be done onto different
simplexes. In this package we present tools to help compare models
(mostly through similations), and this tool compares these objects after
projecting them into a one-dimension less space through the simplexical
structure of the data.

In \pkg{EpiCompare} we allow for the compare epidemics in these higher
dimensional spaces by first projecting onto these simplexes. Even though
higher dimensional models may not be able to visualized, we provide
multiple tools to aid in the comparison of models and epidemics. The
first of which uses multiple simulations under specific model parameters
to assess the bairabilty of the model fit. In \pkg{EpiCompare} we
provide ways to create prediction regions for a true epidemic under the
model assumptions using these simulations. These regions require
representing multi-dimensional structures for functions to completely
contain epidemics, and treat these simulations and epidmeics as
filamental objects. We extend off of papers like \citet{Dalmasso2019a}
to create these bands. These high dimensional bands allow the user to
assess if the true epidemic is within the band (there-by assessing the
model's representation of the epidemic), and also compare who different
models are from each other through distances that compare sets. We
recommend using the Hausdorff distance to compare such sets as it
captures how much bigger the sets would have to expand to cover each
other, and is defined mathematically as \[
d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y) \;,\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
\]

@@

{\centering \includegraphics{Figs/unnamed-chunk-8-1}

@@

\caption{\label{fig:cifs}Empirical cumulative incidence functions of prodrome (symptom) onset and measles rash appearance.  We see that there is approximately a a constant lag between the two curves.}\label{fig:unnamed-chunk-8}
\end{figure}
\end{CodeChunk}

@@

@@

\caption{\label{fig:hag-tern-raw}Time invariant view of the Hagelloch epidemic where we view the individuals in Susceptible, Infectious, or Recovered states.  We see there are two peaks of infection (the vertical axis).}\label{fig:unnamed-chunk-10}
\end{figure}
\end{CodeChunk}

{\centering \includegraphics{Figs/unnamed-chunk-11-1}

\caption{\label{fig:tern-class-data}Time invariant outbreak curves for the three class groups.  The pre-school class has a distinct peak of infection whereas the peak infection point for the other two classes are less well defined.}\label{fig:unnamed-chunk-11}
\end{figure}
\end{CodeChunk}

@@

{\centering \includegraphics{Figs/unnamed-chunk-18-1}

@@

\caption{\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models.}\label{fig:unnamed-chunk-18}
\end{figure}
\end{CodeChunk}

>>>>

The recent (and on-going) COVID-19 global pandemic has galvanized public
interest in understanding more about infectious disease modeling and has
highlighted the usefulness of research in the area of infectious disease
epidemiology. Infectious diseases inflict enormous burdens on the world:
millions of lives lost and trillions of dollars spent yearly. Infectious
disease models typically attempt to do one or more of the following: 1)
predict the spread of current and future epidemics
\citep[e.g. flu prediction][]{Biggerstaff2016}, 2) analyze past and
current epidemics to increase scientific knowledge
\citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or
project epidemic scenarios under pre-specified parameters
\citep[e.g. ...][]{}\footnote{@Shannon: CITE}. At the same time,
descriptive statistics and visualizations (from universities like John
Hopkins \citep{}\footnote{@Shannon: CITE}, many branches and levels of
government, and new organizations \citep{}\footnote{@Shannon: CITE}) are
an important first step of the process.

@@

Although epidemics are by definition a process that evolves over time,
epidemics, including the recent COVID-19, often need to be compared in a
time-invariant way to understand the processes at play. Additionally,
many tools to examine the quantity of the population in each state along
the infection process (for example: quantity of susceptible vs infected
vs recovered individuals) don't always as intelligently capture the
natural connections between the proportion of individuals in these
states. Tools in \pkg{EpiCompare} attempt to give the user the ability
to extend their toolkit to evaluate epidemics to also include
time-invariant approaches and in this section we present benefits of the
time-invariant analysis, with 1) motivation for time-invariant analysis
through \(R_0\) (the initial reproduction number), 2) time-invariant
visualization tools for 3-state models (e.g.~SIR models), and 3)
potential for similar analysis for models with more states (e.g.~SEIR
models).
\textcolor{violet}{Will need to rewrite this section "introduction".}

@@

By definition, \(R_0\) is the number of expected secondary infections
when a primary infection is introduced to a susceptible population.
\(R_0\) is also, maybe, the most famous \textit{time-invariant}
numerical summary of an epidemic, which allows epidemics to be compared
to one another in both different time and geographic scales. For
example, \(R_0\) for Covid-19 is estimated to be between 2-3, seasonal
influenza between 1.2-2, and modern measles outbreaks between 5-6.

@@

With regards to traditional epidemic \(state\) vs.~\(time\) plots,
\(R_0\) is difficult to visualize, especially with respect from one
epidemic to another. For example, consider the scenarios where the first
epidemic is generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
\(\beta_1 = 0.3\) and \(\gamma_1 = 0.15\), and the second epidemic is
generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
\(\beta_2 = 0.24\) and \(\gamma_1 = 0.12\) over 40 days. Both epidemics
have the same value of
\(R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2\). The epidemic
trajectories are shown in the \(state\) vs.~time plots in Figure
\ref{fig:sir-ex}. At a glance, we may assume that Model 1 has a larger
\(R_0\) than Model 2 because the peak of infection occurs more quickly
than in Model 2. On the other hand, we may think Model 2 has a larger
\(R_0\) because it has a slightly larger peak of infection than Model 1.

@@

\caption{\label{fig:different-scales-standard}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.}\label{fig:unnamed-chunk-2}
\end{figure}
\end{CodeChunk}

@@

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
result from \cite{Harko2014} and is shown in detail in Appendix
\ref{app:proof}. The consequence of Theorem \ref{thm:sir-scale} is that
for two SIR models that have the same initial percent of individuals in
each state and \(R_0\) then for every point on the epidemic path of the
first SIR model is also a point on the epidemic path of the second SIR
model. Taking the sample simulations from Fig.
\ref{fig:different-scales-standard}, Fig.
\ref{fig:different-scales-tern} presents these two models in a ternary
plot. This means, that at a glance, we can tell if two epidemics have
different values of \(R_0\).

@@

\caption{\label{fig:different-scales-tern}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.  These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered.}\label{fig:unnamed-chunk-3}
\end{figure}
\end{CodeChunk}

\hypertarget{beyond-the-kermack-and-mckendrick-sir-models}{%
\subsection{Beyond the Kermack and McKendrick SIR
models}\label{beyond-the-kermack-and-mckendrick-sir-models}}

Although the result of Theorem \ref{thm:sir-scale} allows for easy
visual comparison of \(R_0\) in Kermack SIR models, it does require
stringent assumptions of homogeneity of behavior in populations. The use
of visualizing epidemics in a time-invariant lens via ternary plots
extends beyond those of models that follow the ODEs in the
Kermack-McKendrick equations. Any model with S, I, and R states can be
visualized with ternary plots, regardless of birth and death dynamics
and regardless of homogeneity of individual behavior. We can use ternary
plots to compare the spread of a disease for groups within a population
without time as a confounding factor. We show an example of this in a
later section.

But time-invariant analysis is also useful for epidemic models with more
than three states. The constraints in three dimensions that are met with
the SIR model (that is
\(\sum_{i=1}^3 (\text{number in state}(i)) = N(t)\)) actually represents
a space of 3d simplices, and the ternary plot specifically represents
these simplices, after scaling to examine such values as proportions
between 0 and 1 (ternary plots are known as a 3d unit simplex due to
it's scaling). This same scaling for larger models (i.e.~with more
states) can be done onto different simplexes. In this package we present
tools to help compare models (mostly through simulations). s tool
compares these objects after projecting them into a one-dimension-fewer
space through the simplexical structure of the data.

In \pkg{EpiCompare} we allow for the comparison epidemics in these
higher dimensional spaces by first projecting onto these simplexes. Even
though higher dimensional models may not be able to visualized, we
provide multiple tools to aid in the comparison of models and epidemics.
The first of which uses multiple simulations under specific model
parameters to assess the bairabilty of the model fit. In
\pkg{EpiCompare} we provide ways to create prediction regions for a true
epidemic under the model assumptions using these simulations. These
regions require representing multi-dimensional structures for functions
to completely contain epidemics, and treat these simulations and
epidmeics as filamental objects. We extend off of papers like
\citet{Dalmasso2019a} to create these bands. These high dimensional
bands allow the user to assess if the true epidemic is within the band
(there-by assessing the model's representation of the epidemic), and
also compare who different models are from each other through distances
that compare sets. We recommend using the Hausdorff distance to compare
such sets as it captures how much bigger the sets would have to expand
to cover each other, and is defined mathematically as \[
d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y) \;,\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
\]

@@

{\centering \includegraphics{Figs/unnamed-chunk-7-1}

@@

\caption{\label{fig:cifs}Empirical cumulative incidence functions of prodrome (symptom) onset and measles rash appearance.  We see that there is approximately a a constant lag between the two curves.}\label{fig:unnamed-chunk-7}
\end{figure}
\end{CodeChunk}

@@

{\centering \includegraphics{Figs/unnamed-chunk-9-1}

\caption{\label{fig:hag-tern-raw}Time invariant view of the Hagelloch epidemic where we view the individuals in Susceptible, Infectious, or Recovered states.  We see there are two peaks of infection (the vertical axis).}\label{fig:unnamed-chunk-9}
\end{figure}
\end{CodeChunk}

@@

\caption{\label{fig:tern-class-data}Time invariant outbreak curves for the three class groups.  The pre-school class has a distinct peak of infection whereas the peak infection point for the other two classes are less well defined.}\label{fig:unnamed-chunk-10}
\end{figure}
\end{CodeChunk}

@@

{\centering \includegraphics{Figs/unnamed-chunk-17-1}

@@

\caption{\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models.}\label{fig:unnamed-chunk-17}
\end{figure}
\end{CodeChunk}
